---
title: "AE 03 - Visualizing Star Wars Characters"
author: "Jaffer Ali"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
  pdf_document:
    toc: true
    latex_engine: xelatex
    keep_tex: false
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## Introduction

In this application exercise, you'll practice creating various types of visualizations using the Star Wars characters dataset. This dataset contains information about characters from the Star Wars universe, including their physical characteristics, homeworld, and species.

**In this application exercise, you will:**

- Explore data with `glimpse()`
- Create scatterplots with multiple aesthetics
- Make histograms with appropriate binwidths
- Create visualizations for different variable types
- Interpret relationships between variables
- Customize plot labels and aesthetics
- Choose appropriate plot types for different data types

**Estimated time:** 30-45 minutes

---

## Getting Started

### Fork and Clone Workflow

**Quick reminder:** You should be working in YOUR forked `application-exercises` repository.

1. Navigate to the `ae-03-starwars` folder in YOUR fork
2. Open `ae-03-starwars.Rmd`
3. Update the YAML with your name
4. Click **Knit** to verify it works

**If you need detailed fork-and-clone instructions, see AE 01.**

---

## Packages
```{r load-packages, include=FALSE}
library(tidyverse)
```

**What this package does:**
- **tidyverse:** Includes ggplot2 for visualization and dplyr for data wrangling

---

## The Data

The `starwars` dataset comes built-in with the tidyverse. It contains information on Star Wars characters from the first seven films.

---

## Exercise 1: Glimpse at the Data

Glimpse at the `starwars` data frame to understand its structure.
```{r glimpse-starwars}
glimpse(starwars)
```

**Answer the following questions:**

**How many observations (rows) are there?** 87

**How many variables (columns) are there?** 14

**Name three numerical variables:**

1. Height 
2. Mass
3. Birth year

**Name three categorical variables:**

1. hair_color 
2.skin_color 
3. eye_color


**What does each observation represent?**

Each observation represents one individual character,including their physical characteristics, demographic information, and related attributes.

## Exercise 2: Understanding Aesthetics

The plot below maps several variables to aesthetics. However, there's an issue with how color is being used.

**Original code:**
```{r scatterplot-original, eval=FALSE}
ggplot(starwars, 
       aes(x = height, y = mass, color = gender, size = birth_year)) +
  geom_point(color = "pink")
```

**Task 2.1:** Run the code above. What happened to the `color = gender` mapping?

**Your answer:**

The color mapping to gender was overridden, so all points appear the same color instead of being colored by gender. The plot now only shows the mapping for birth_year (size), and the gender color legend disappeared.

**Task 2.2:** Why did this happen?

**Hint:** Think about where `color = "pink"` appears vs. where `color = gender` appears.

**Your answer:**

This happened because the color aesthetic was overridden or removed in the plotting code. As a result, ggplot no longer mapped color to gender, and all points were displayed using a single color instead of different colors for each gender.

**The issue:** When you set `color = "pink"` inside `geom_point()`, it overrides the aesthetic mapping `color = gender` from `aes()`.

**Task 2.3:** Modify the code below to keep all the aesthetic mappings AND make the points pink:
```{r scatterplot-pink}
ggplot(starwars, 
       aes(x = height, y = mass, size = birth_year)) +  # Removed color = gender
  geom_point(color = "pink")
```

**What this plot shows:**

- x-axis: Height of characters
- y-axis: Mass (weight) of characters  
- Size of points: Birth year (larger = born later)
- All points are pink

---

## Exercise 3: Adding Labels

Add appropriate labels for the title, x-axis, y-axis, and size legend. Uncomment the lines and fill in the blanks.
```{r scatterplot-labels}
ggplot(starwars, 
       aes(x = height, y = mass, color = gender, size = birth_year)) +
  geom_point(color = "#30509C") +
  labs(
    title ="Mass vs Height of Star Wars Charchters",           # Title about relationship between height and mass
    x = "Height(cm)",               # What is on the x-axis?
    y = "Mass(kg)",               # What is on the y-axis?
    size = "Birth Year"             # What does size represent?
  )
```

**Hints:**

- Title should describe what the plot shows
- x and y labels should include units if relevant (height in cm, mass in kg)
- size label describes what the size of points represents

**Task 3.1:** Why is there no `color` label in the `labs()` function?

**Your answer:**

There is no color label because the color mapping to gender was overridden by setting a fixed color in geom_point(). Since color is no longer mapped to a variable, ggplot does not create a color legend, and therefore no color label is needed in labs().
---

## Exercise 4: Create a Histogram

Pick a single numerical variable and make a histogram of it. Select a reasonable binwidth.

**First, explore the variable:**
```{r explore-for-histogram}
# Choose a variable and look at its range
starwars %>%
  summarise(
    min_height = min(height, na.rm = TRUE),
    max_height = max(height, na.rm = TRUE),
    mean_height = mean(height, na.rm = TRUE)
  )
```

**Now create your histogram:**
```{r histogram, eval = TRUE}
# Fill in the blanks
ggplot(starwars, aes(x = height)) +
  geom_histogram(binwidth = 10, fill = "steelblue", color = "white") +
  labs(
    title = "Distribuation of Star Wars Charchter Heights",
    x = "Heights(cm)",
    y = "Number of Charhcters"
  )
```

**Hints:**

- Choose from: height, mass, birth_year
- Try different binwidths: 10, 20, 50 (depending on your variable)
- Use `fill` to color the bars, `color` to color the borders

**Task 4.1:** What binwidth did you choose and why?

**Your answer:**

I chose a binwidth of 10 because it creates bins that are wide enough to show the overall pattern of the data while still providing enough detail to understand the distribution of character heights.

**Task 4.2:** Describe the distribution. Is it symmetric, skewed left, or skewed right? Are there any outliers?

**Your answer:**

The distribution is slightly skewed right with a few tall outliers. Most characters cluster around the middle height range.

---

## Exercise 5: Numerical vs. Categorical

Pick a numerical variable and a categorical variable and make a visualization to visualize the relationship between the two variables.

**Suggested plot types:**

- Boxplot: `geom_boxplot()`
- Violin plot: `geom_violin()`
- Multiple histograms: `geom_histogram()` with `facet_wrap()`

**Example combinations to try:**

- height by gender
- mass by species
- birth_year by sex
```{r num-cat, eval = TRUE}
# Your code here
ggplot(starwars, aes(x = gender, y = height)) +
  geom_boxplot(fill = "lightblue") +
  labs(
    title = "Height by Gender of Star Wars Characters",
    x = "Gender",
    y = "Height (cm)"
  )
```

**Interpretation:**

What does this visualization tell you about the relationship between these two variables?

This visualization shows how the heights of Star Wars characters vary by gender. The boxplot allows us to compare the distribution of heights for each gender category. Most of the characters identified as masculine appear to have a wider range of heights, while other gender categories have fewer observations and less variation. Overall, the plot suggests that height differs somewhat between gender groups, but the masculine category contains most of the data points in the dataset.

---

## Exercise 6: Bar Plot of Categorical Variable

Pick a single categorical variable from the dataset and make a bar plot of its distribution.

**Suggested variables:**

- gender
- sex  
- species
- eye_color
- hair_color
```{r barplot,, eval = TRUE}
# Your code here
ggplot(starwars, aes(x = gender)) +
  geom_bar(fill = "coral") +
  labs(
    title = "Distribution of Star Wars Characters by Gender",
    x = "Gender",
    y = "Number of Characters"
  )
```

**Task 6.1:** Which category has the most observations?

**Your answer:** Masculine 

**Task 6.2:** If your plot has many categories with small bars that are hard to read, try flipping the coordinates:
```{r barplot-flipped, eval = TRUE}
# Same code as above, but add coord_flip() at the end
ggplot(starwars, aes(x = gender)) +
  geom_bar(fill = "coral") +
  labs(
    title = "Distribution of Star Wars Characters by Gender",
    x = "Gender",
    y = "Number of Characters"
  ) +
  coord_flip()
```

---

## Exercise 7: Two Categorical Variables

Pick two categorical variables and make a visualization to visualize the relationship between the two variables.

**Suggested plot types:**

- Stacked bar plot: `geom_bar()`
- Grouped bar plot: `geom_bar(position = "dodge")`
- Faceted bar plot: `geom_bar()` with `facet_wrap()`
- Heatmap: `geom_tile()` with counts

**Suggested combinations:**

- gender and species
- sex and eye_color
- gender and hair_color

**Option 1: Stacked bar plot**
```{r cat-cat-stacked, eval = TRUE}
# Your code here
ggplot(starwars, aes(x = gender, fill = species)) +
  geom_bar() +
  labs(
    title = "Species Distribution by Gender",
    x = "Gender",
    y = "Count",
    fill = "Species"
  )
```

**Option 2: Grouped bar plot (side-by-side)**
```{r cat-cat-grouped, eval = FALSE}
# Your code here
ggplot(starwars, aes(x = ___, fill = ___)) +
  geom_bar(position = "dodge") +
  labs(
    title = "___",
    x = "___",
    y = "Count",
    fill = "___"
  )
```

**Choose one of the above and interpret:**

What does this visualization tell you about the relationship between these two variables?

This plot shows that most Star Wars characters are masculine, and they belong to many different species. The feminine group has fewer characters, and the NA group has very few.

---

## Exercise 8: Multi-Variable Visualization

Pick two numerical variables and two categorical variables and make a visualization that incorporates all of them.

**Suggested approach:**

- x-axis: numerical variable 1
- y-axis: numerical variable 2
- color: categorical variable 1
- facets: categorical variable 2

**Example combination:**

- height (numerical)
- mass (numerical)
- gender (categorical - use for color)
- species (categorical - use for facets, maybe filter to top 3-4 species)

**Starter template:**
```{r multi, eval=FALSE}
# You might want to filter first to avoid too many facets
starwars_filtered <- starwars %>%
  filter(species %in% c("Human", "Droid", "Wookiee"))  # Pick 3-4 species

# Your plot here
ggplot(starwars_filtered, 
       aes(x = height, y = mass, color = gender)) +
  geom_point(size = 3) +
  facet_wrap(~ species) +
  labs(
    title = "Height vs Mass of Star Wars Characters by Species",
    x = "Height (cm)",
    y = "Mass (kg)",
    color = "Gender"
  )
```

**Interpretation:**

What patterns do you notice? How do the relationships differ across the facets?

The plot shows that taller characters usually have higher mass. Humans have the most characters and a wider range of heights and masses. Wookiees are very tall and heavy, while droids have fewer characters with different sizes.

---

## Exercise 9: Reflection

**Task 9.1:** Which type of plot was most useful for exploring this data? Why?

**Your answer:**

The scatter plot with facets was the most useful because it shows the relationship between height and mass while also comparing different species and genders. This makes it easier to see patterns and differences between groups.

**Task 9.2:** What was the most interesting pattern or relationship you found?

**Your answer:**

The most interesting pattern was that Wookiees are much taller and heavier than most other characters, while humans show a wider range of heights and masses.

**Task 9.3:** What did you find most challenging about creating these visualizations?

**Your answer:**

The most challenging part was choosing the right variables and plot types to clearly show the relationships in the data. Some variables also had many categories, which made the plots harder to read.

---

## Knit and Submit

**Before you finish:**

1. ✅ Make sure all code chunks run without errors
2. ✅ Fill in all answer spaces and interpretations
3. ✅ Knit to HTML first to verify everything works
4. ✅ Review the HTML output - check that all plots display correctly

**Git workflow:**

1. Click the **Git** pane
2. Check boxes next to all changed files
3. Click **Commit**
4. Write a commit message: "Completed AE 03 - Star Wars"
5. Click **Commit**
6. Click **Push**
7. Verify your work is on GitHub in **YOUR fork**

---

## Generate PDF for Canvas Submission

Once everything works in HTML:

1. Click the **Knit** dropdown arrow
2. Select **Knit to PDF**
3. Wait for PDF generation
4. Find `ae-03-starwars.pdf` in the Files pane
5. Export and save to your computer
6. Submit the PDF to Canvas

**Remember:** Submit to BOTH GitHub (your .Rmd) AND Canvas (the PDF)!

---

## Troubleshooting

**Common issues:**

- **Error: "object not found":** Make sure you loaded the tidyverse package
- **Plot doesn't show up:** Make sure the code chunk is set to run (not `eval: false`)
- **Too many categories in plot:** Filter the data first or use `coord_flip()` for better readability
- **NAs in plot:** This is normal for this dataset - some characters have missing values

**Aesthetic mapping issues:**

- **Color not showing:** Make sure `color = variable` is inside `aes()`, not `geom_point()`
- **Want to set all points to one color:** Put `color = "colorname"` OUTSIDE `aes()`, inside `geom_point()`

**Plot type selection:**

- **One numerical variable:** histogram or density plot
- **One categorical variable:** bar plot
- **Numerical vs. categorical:** boxplot, violin plot, or faceted histogram
- **Two categorical:** stacked/grouped bar plot or faceted plot
- **Two numerical:** scatterplot
- **Multiple variables:** use color, size, shape, facets

**PDF issues:**

- **LaTeX not found:** Install tinytex: `tinytex::install_tinytex()`
- **Plots too large for PDF:** Adjust `fig.width` and `fig.height` in chunk options

**If you're stuck:**

- Review the walkthrough video
- Check ggplot2 documentation: `?geom_point`, `?geom_histogram`, etc.
- Ask on the discussion board
- Attend office hours

---

## Bonus: Quick Reference

**Basic ggplot structure:**
```r
ggplot(data, aes(x = var1, y = var2)) +
  geom_*() +
  labs(title = "...", x = "...", y = "...")
```

**Common geoms:**

- `geom_point()`: scatterplot
- `geom_histogram()`: histogram
- `geom_bar()`: bar plot
- `geom_boxplot()`: boxplot
- `geom_violin()`: violin plot
- `geom_line()`: line plot

**Common aesthetics (inside `aes()`):**

- `x = variable`: x-axis
- `y = variable`: y-axis
- `color = variable`: color by variable
- `fill = variable`: fill color by variable
- `size = variable`: size by variable
- `shape = variable`: shape by variable

**Fixed aesthetics (outside `aes()`):**

- `color = "blue"`: all points blue
- `fill = "red"`: all filled red
- `size = 3`: all points size 3
- `alpha = 0.5`: transparency (0 = transparent, 1 = opaque)